Tropical Mathematics and the Lambda-calculus I

### 2023-11-30 CSL24/section1.tex
commit

On the one hand, one one could mention the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask   whether they compute functions which do not differ \emph{too much}. This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in, e.g., differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.

>>>>

On the one hand, there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask   whether they compute functions which do not differ \emph{too much}. This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in, e.g., differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.

### 2023-11-30 CSL24/section2bis.tex
Commit notturno
commit

At the basis of our approach is the observation that the \emph{tropical semiring} $([0,\infty], \min, +)$, which is at the heart of tropical mathematics, coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$ \cite{Hofmann2014, Stubbe2014}, the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{Lawvere1973}.
Let us recall that a quantale is a complete lattice endowed with a continuous monoid action. In the case of $\Lawv$ the lattice is defined by the reverse order $\geq$ on $\BB R$, and the monoid action is provided by addition. Notice that the lattice join operation of $\Lawv$ coincides with the idempotent semiring operation $\min$.

@@

\begin{definition}
A \emph{tropical power series} (tps) 
%\footnote{In analogy with \cite{DanEhrh2011}, we could also call them \emph{tropical entire functions}.} 
%(tps) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$.
in $k$-variables is a function $f:\Lawv^k\to\Lawv$ of shape $f(x)=\inf_{i\in I}\{ix + \matr f(i) \}$, where 
$I\subseteq\mathbb N^{k}$, $i x$ is the scalar product and $\matr f\in \Lawv^{{\mathbb N}^{k}}$ is a vector of coefficients.
When $I$ is finite, $f$ is called a \emph{tropical polynomial}. 
\end{definition}
Hence, a unary tps is a function $f:\Lawv\to \Lawv$ of the form $f(x)=\inf_{i\in I}\{ix+a_{i}\}$, with $I\subseteq\mathbb N$ and the $a_{i}\in \Lawv$.
In Section \ref{section3} we also consider tps in \emph{infinitely many} variables.

@@

%For instance, the tps $\varphi(x,y)=\inf_{n}\{2x+(n+1)y,3x\}$ coincides with the polynomial $\min\{2x+y,3x\}$, since $2x+(n+1)y\geq 2x+y$ holds for all $n$. So the trivial valuation sends both the infinite series $\sum_{n}(x^{2}y^{n+1}+x^{3})$ and the polynomial $x^{2}y+x^{3}$ onto the same map.
Notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$ and the zero is $\infty$.
Therefore, for instance, the tropicalisation (w.r.t.\ any valuation) sends the infinite power series $\sum_{n\geq 2} x^n$ to the tps $\varphi(x)=\inf_n (n+2)x$.
But the latter always coincides with the tropical polynomial $\varphi(x)=2x$, since $nx\geq 0$ for all $n\in\N$, which is in turn the tropicalisation of the polynomial $x^2$.
This shows that tropicalisation is \emph{not} in general an injective operation and in fact, as we show in \autoref{theorem:fepsilon} below, tps (in finitely many variables) have a tendency to collapse, if not globally at least locally, onto tropical polynomials.

@@

\begin{remark}\label{rmk:AC}
If $U\subseteq \N^{K}$ is infinite, then $U$ contains an infinite ascending chain $ m  _{0}\prec  m  _{1} \prec  m  _{2} \prec \dots$.
This is a consequence of K\"onig Lemma (KL): consider the directed acyclic graph $(U,\prec_{1})$, indeed a $K$-branching tree; if there is no infinite ascending chain $  m  _{0}\prec  m  _{1} \prec  m  _{2} \prec \dots$, then in particular there is no infinite ascending chain $  m  _{0}\prec_{1}  m  _{1} \prec_{1}  m  _{2} \prec_{1} \dots$ so the tree $U$ has no infinite ascending chain; then by KL it is finite, contradicting the assumption. 
\end{remark}

%Theorem \ref{theorem:fepsilon} now follows from the following result:
%\begin{proposition}%[Theorem \ref{theorem:fepsilon}]
%Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tps with matrix $\matr f:\N^k\to\Lawv$.
% For all $0<\epsilon<\infty$, there is $\C F_\epsilon \subseteq_{\mathrm{fin}} \N^k$ such that:
%\begin{enumerate}
%% \item $\C F_\epsilon$ is finite
% \item if $\mathcal{F}_\epsilon= \emptyset$ then $f( x ) = +\infty$ for all $ x \in \Lawv^k$;
% \item if $f( x _0) = +\infty$ for some $ x _0\in [\epsilon,\infty)^{K}$ then $\mathcal{F}_\epsilon= \emptyset$;
% \item the restriction of $f$ on $[\epsilon,\infty]^k$ coincides  with  $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{n x+\matr f(n)}$.
%%where $nx:=\sum_{i=1}^k n_ix_i$.
%\end{enumerate}
%\end{proposition}
\begin{proof}[Proof of Theorem \ref{theorem:fepsilon}]
We will actually show the existence of $\C F_\epsilon \subseteq_{\mathrm{fin}} \N^k$ such that:
\begin{enumerate}
% \item $\C F_\epsilon$ is finite
 \item if $\mathcal{F}_\epsilon= \emptyset$ then $f( x ) = +\infty$ for all $ x \in \Lawv^k$;
 \item if $f( x _0) = +\infty$ for some $ x _0\in [\epsilon,\infty)^{K}$ then $\mathcal{F}_\epsilon= \emptyset$;
 \item the restriction of $f$ on $[\epsilon,\infty]^k$ coincides  with  $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{n x+\matr f(n)}$.
%where $nx:=\sum_{i=1}^k n_ix_i$.
\end{enumerate}
Let $\mathcal F_\epsilon$ be the complementary in $\N$ of the set:
\[
 \set{ n  \in\N^{K} \mid \textit{either } \hat f ( n  )=+\infty \textit{ or there is }  m  \prec  n  \textit{ s.t.\ } \hat f( m  )\leq\hat f( n  )+\epsilon}.
\]
In other words, $ n  \in\mathcal F_\epsilon$ iff $\hat f( n  )<+\infty$ and for all $ m  \prec  n  $, one has $\hat f( m  )>\hat f( n  )+\epsilon$.

Suppose that $\mathcal F_\epsilon$ is infinite; then, using Remark~\ref{rmk:AC}, it contains an infinite ascending chain $\set{ m  _0\prec  m  _1\prec\cdots}$.  
By definition of $\mathcal F_\epsilon$ we have then 
$+\infty>\hat f( m  _0)>\hat f( m  _1)+\epsilon>\hat f( m  _2)+2\epsilon>\cdots$, 
so that $+\infty>\hat f( m  _0)>\hat f( m  _{i})+i\epsilon\geq i\epsilon$ for all $i\in\N$.
This contradicts the Archimedean property of $\R$.

1).
We show that if $\mathcal F_\epsilon=\emptyset$, then $\hat f( n  )=+\infty$ for all $ n  \in\N^{K}$.
This immediately entails the desired result.
We go by induction on the well-founded order $\prec$ over $ n  \in\N^{K}$:

- if $ n  =0^{K}\notin\mathcal F_\epsilon$, then $\hat f( n  )=+\infty$, because there is no $ m  \prec n  $.

- if $ n  \notin\mathcal F_\epsilon$, with $ n  \neq 0^{K}$ then either $\hat f( n  )=+\infty$ and we are done, or there is $ m  \prec  n  $ s.t.\ $\hat f( m  )\leq \hat f( n  )+\epsilon$.
By induction $\hat f( m  )=+\infty$ and, since $\epsilon<+\infty$, this entails $\hat f( n  )=+\infty$.

2).
If $f( x _0)=+\infty$ with $ x _0\in [\epsilon,\infty)^{K}$, then necessarily $\hat f( n  )=+\infty$ for all $ n  \in\N^{K}$.
Therefore, no $ n  \in\N^{K}$ belongs to $\mathcal F_\epsilon$.

3).
We have to show that $f( x )=P_\epsilon( x )$ for all $ x \in [\epsilon,+\infty]^{K}$.
By 1), it suffices to show that we can compute $f( x )$ by taking the $\inf$, that is therefore a $\min$, only in $\mathcal F_\epsilon$ (instead of all $\N^{K}$).
If $\mathcal F_\epsilon=\emptyset$ then by 2) we are done (remember that $\min\emptyset := +\infty$).
If $\mathcal F_\epsilon\neq\emptyset$, we show that for all $ n  \in\N^{K}$, if $ n   \notin\mathcal F_\epsilon$, then there is $ m  \in\mathcal F_\epsilon$ s.t.\ $\hat f( m  )+ m   x  \leq \hat f( n  )+ n   x $.
We do it again by induction on $\prec_{1}$:

- if $ n  =0^{K}$, then from $\mathbf  n\notin \mathcal F_{\epsilon}$, by definition of $\mathcal F_\epsilon$, we have $\hat f( n  )=+\infty$ (because there is no $ n  '\prec n  $).
So any element of $\mathcal F_\epsilon\neq\emptyset$ works.

- if $ n  \neq 0^{K}$, then we have two cases:
either $\hat f( n  )=+\infty$, in which case we are done as before by taking any element of $\mathcal F_\epsilon\neq\emptyset$.
Or $\hat f( n  )<+\infty$, in which case (again by definition of $\mathcal F_\epsilon$) there is $ n  '\prec n  $ such that $ \hat f( n  ')\leq \hat f( n  )+\epsilon$  ($\star$).
Therefore we have (remark that the following inequalities hold also for the case $x=+\infty$):
\[\begin{array}{rclr}
 \hat f( n  ')+ n  ' x  & \leq & \hat f( n  ) + \epsilon +  n' x  & \textit{by }(\star) \\
 & \leq & \hat f( n  ) + ( n  - n  ') x  +  n  ' x  & \textit{because $\epsilon\leq\min x $ and $\min  x \leq( n  -  n') x $} \\
 & = & \hat f( n  )+  n   x . &
\end{array}\]
Now, if $ n  '\in\mathcal F_\epsilon$ we are done.
Otherwise $ n  '\notin\mathcal F_\epsilon$ and we can apply the induction hypothesis on it, obtaining an $ m  \in\mathcal F_\epsilon$ s.t.\ $\hat f( m  )+ m   x  \leq \hat f( n  ')+ n  ' x $.
Therefore this $ m  $ works.
\end{proof}

>>>>

At the basis of our approach is the observation that the \emph{tropical semiring} $([0,\infty], \min, +)$, which is at the heart of tropical mathematics, coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$ \cite{Hofmann2014, Stubbe2014}, the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{Lawvere1973}.
Let us recall that a quantale is a complete lattice endowed with a continuous monoid action.
In the case of $\Lawv$, the lattice is defined by the reverse order $\geq$ on $\BB R$, and the monoid action is provided by addition. Notice that the lattice join operation of $\Lawv$ coincides with the idempotent semiring operation $\min$.

@@

\begin{definition}
A \emph{tropical power series} (tps) 
%\footnote{In analogy with \cite{DanEhrh2011}, we could also call them \emph{tropical entire functions}.} 
%(tps) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$.
in $k$ variables is a function $f:\Lawv^k\to\Lawv$ of shape $f(x)=\inf_{i\in I}\{ix + \matr f(i) \}$, where 
$I\subseteq\mathbb N^{k}$, $i x$ is the scalar product and $\matr f\in \Lawv^{{\mathbb N}^{k}}$ is a vector of coefficients.
When $I$ is finite, $f$ is called a \emph{tropical polynomial}. 
\end{definition}
Hence, a unary tps is a function $f:\Lawv\to \Lawv$ of the form $f(x)=\inf_{i\in I}\{ix+a_{i}\}$, with $I\subseteq\mathbb N$ and the $a_{i}\in \Lawv$.
In Section \ref{section3} we also consider tps in \emph{infinitely many} variables.

@@

%For instance, the tps $\varphi(x,y)=\inf_{n}\{2x+(n+1)y,3x\}$ coincides with the polynomial $\min\{2x+y,3x\}$, since $2x+(n+1)y\geq 2x+y$ holds for all $n$. So the trivial valuation sends both the infinite series $\sum_{n}(x^{2}y^{n+1}+x^{3})$ and the polynomial $x^{2}y+x^{3}$ onto the same map.
Notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$ and the zero is $\infty$.
Therefore, for instance, the tropicalisation (w.r.t.\ any valuation) sends the infinite power series $\sum_{n\geq 2} x^n$ to the tps $\varphi(x)=\inf_n (n+2)x$.
Yet the latter always coincides with the tropical polynomial $\varphi(x)=2x$, since $nx\geq 0$ for all $n\in\N$, which is in turn the tropicalisation of the polynomial $x^2$.
This shows that tropicalisation is \emph{not} in general an injective operation and in fact, as we show in \autoref{theorem:fepsilon} below, tps (in finitely many variables) have a tendency to collapse, if not globally at least locally, onto tropical polynomials.

@@

\begin{remark}\label{rmk:AC2}
If $U\subseteq \N^{K}$ is infinite, then $U$ contains an infinite ascending chain $ m  _{0}\prec  m  _{1} \prec  m  _{2} \prec \dots$.
This is a consequence of K\"onig Lemma (KL): consider the directed acyclic graph $(U,\prec_{1})$, indeed a $K$-branching tree; if there is no infinite ascending chain $  m  _{0}\prec  m  _{1} \prec  m  _{2} \prec \dots$, then in particular there is no infinite ascending chain $  m  _{0}\prec_{1}  m  _{1} \prec_{1}  m  _{2} \prec_{1} \dots$ so the tree $U$ has no infinite ascending chain; then by KL it is finite, contradicting the assumption. 
\end{remark}

%Theorem \ref{theorem:fepsilon} now follows from the following result:
%\begin{proposition}%[Theorem \ref{theorem:fepsilon}]
%Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tps with matrix $\matr f:\N^k\to\Lawv$.
% For all $0<\epsilon<\infty$, there is $\C F_\epsilon \subseteq_{\mathrm{fin}} \N^k$ such that:
%\begin{enumerate}
%% \item $\C F_\epsilon$ is finite
% \item if $\mathcal{F}_\epsilon= \emptyset$ then $f( x ) = +\infty$ for all $ x \in \Lawv^k$;
% \item if $f( x _0) = +\infty$ for some $ x _0\in [\epsilon,\infty)^{K}$ then $\mathcal{F}_\epsilon= \emptyset$;
% \item the restriction of $f$ on $[\epsilon,\infty]^k$ coincides  with  $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{n x+\matr f(n)}$.
%%where $nx:=\sum_{i=1}^k n_ix_i$.
%\end{enumerate}
%\end{proposition}
\begin{proof}[Proof of Theorem \ref{theorem:fepsilon}]
We will actually show the existence of $\C F_\epsilon \subseteq_{\mathrm{fin}} \N^k$ such that:
\begin{enumerate}
% \item $\C F_\epsilon$ is finite
 \item if $\mathcal{F}_\epsilon= \emptyset$ then $f( x ) = +\infty$ for all $ x \in \Lawv^k$;
 \item if $f( x _0) = +\infty$ for some $ x _0\in [\epsilon,\infty)^{K}$ then $\mathcal{F}_\epsilon= \emptyset$;
 \item the restriction of $f$ on $[\epsilon,\infty]^k$ coincides  with  $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{n x+\matr f(n)}$.
%where $nx:=\sum_{i=1}^k n_ix_i$.
\end{enumerate}
Let $\mathcal F_\epsilon$ be the complementary in $\N$ of the set:
\[
 \set{ n  \in\N^{K} \mid \textit{either } \hat f ( n  )=+\infty \textit{ or there is }  m  \prec  n  \textit{ s.t.\ } \hat f( m  )\leq\hat f( n  )+\epsilon}.
\]
In other words, $ n  \in\mathcal F_\epsilon$ iff $\hat f( n  )<+\infty$ and for all $ m  \prec  n  $, one has $\hat f( m  )>\hat f( n  )+\epsilon$.
Suppose that $\mathcal F_\epsilon$ is infinite; then, using Remark~\ref{rmk:AC2}, it contains an infinite ascending chain $\set{ m  _0\prec  m  _1\prec\cdots}$.  
By definition of $\mathcal F_\epsilon$ we have then 
$+\infty>\hat f( m  _0)>\hat f( m  _1)+\epsilon>\hat f( m  _2)+2\epsilon>\cdots$, 
so that $+\infty>\hat f( m  _0)>\hat f( m  _{i})+i\epsilon\geq i\epsilon$ for all $i\in\N$.
This contradicts the Archimedean property of $\R$. Hence $\mathcal F_{\epsilon}$ is finite.

\begin{enumerate}
\item
We show that if $\mathcal F_\epsilon=\emptyset$, then $\hat f( n  )=+\infty$ for all $ n  \in\N^{K}$.
This immediately entails the desired result.
We go by induction on the well-founded order $\prec$ over $ n  \in\N^{K}$:
\begin{itemize}
\item if $ n  =0^{K}\notin\mathcal F_\epsilon$, then $\hat f( n  )=+\infty$, because there is no $ m  \prec n  $.

\item if $ n  \notin\mathcal F_\epsilon$, with $ n  \neq 0^{K}$ then either $\hat f( n  )=+\infty$ and we are done, or there is $ m  \prec  n  $ s.t.\ $\hat f( m  )\leq \hat f( n  )+\epsilon$.
By induction $\hat f( m  )=+\infty$ and, since $\epsilon<+\infty$, this entails $\hat f( n  )=+\infty$.
\end{itemize}
\item If $f( x _0)=+\infty$ with $ x _0\in [\epsilon,\infty)^{K}$, then necessarily $\hat f( n  )=+\infty$ for all $ n  \in\N^{K}$.
Therefore, no $ n  \in\N^{K}$ belongs to $\mathcal F_\epsilon$.

\item We have to show that $f( x )=P_\epsilon( x )$ for all $ x \in [\epsilon,+\infty]^{K}$.
By 1), it suffices to show that we can compute $f( x )$ by taking the $\inf$, that is therefore a $\min$, only in $\mathcal F_\epsilon$ (instead of all $\N^{K}$).
If $\mathcal F_\epsilon=\emptyset$ then by 2) we are done (remember that $\min\emptyset := +\infty$).
If $\mathcal F_\epsilon\neq\emptyset$, we show that for all $ n  \in\N^{K}$, if $ n   \notin\mathcal F_\epsilon$, then there is $ m  \in\mathcal F_\epsilon$ s.t.\ $\hat f( m  )+ m   x  \leq \hat f( n  )+ n   x $.
We do it again by induction on $\prec_{1}$:
\begin{itemize}
\item if $ n  =0^{K}$, then from $\mathbf  n\notin \mathcal F_{\epsilon}$, by definition of $\mathcal F_\epsilon$, we have $\hat f( n  )=+\infty$ (because there is no $ n  '\prec n  $).
So any element of $\mathcal F_\epsilon\neq\emptyset$ works.

\item if $ n  \neq 0^{K}$, then we have two cases:
either $\hat f( n  )=+\infty$, in which case we are done as before by taking any element of $\mathcal F_\epsilon\neq\emptyset$.
Or $\hat f( n  )<+\infty$, in which case (again by definition of $\mathcal F_\epsilon$) there is $ n  '\prec n  $ such that $ \hat f( n  ')\leq \hat f( n  )+\epsilon$  ($\star$).
Therefore we have (remark that the following inequalities hold also for the case $x=+\infty$):
\[\begin{array}{rclr}
 \hat f( n  ')+ n  ' x  & \leq & \hat f( n  ) + \epsilon +  n' x  & \textit{by }(\star) \\
 & \leq & \hat f( n  ) + ( n  - n  ') x  +  n  ' x  & \textit{because $\epsilon\leq\min x $ and $\min  x \leq( n  -  n') x $} \\
 & = & \hat f( n  )+  n   x . &
\end{array}\]
Now, if $ n  '\in\mathcal F_\epsilon$ we are done.
Otherwise $ n  '\notin\mathcal F_\epsilon$ and we can apply the induction hypothesis on it, obtaining an $ m  \in\mathcal F_\epsilon$ s.t.\ $\hat f( m  )+ m   x  \leq \hat f( n  ')+ n  ' x $.
Therefore this $ m  $ works.
\end{itemize}
\end{enumerate}
\end{proof}
